---
title: "Longitudinal Analysis: From Observation to Policy Insight"
author: "Dan Swart"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Longitudinal Analysis: From Observation to Policy Insight}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
library(policycraft)
```

## The analytical purpose

Policy analysts often receive a series of measurements and are asked whether
conditions are improving, worsening, or responding to a policy change. A table
of averages or a comparison between two selected years can conceal the order in
which observations occurred. Longitudinal analysis begins by preserving that
order.

`policycraft` organizes the work as **Observe → Reason → Insight**:

- **Observe:** establish what was measured, how, when, and under which system.
- **Reason:** examine temporal patterns, predictable limits, runs, direction,
  data quality, and plausible alternative explanations.
- **Insight:** explain what the evidence supports, what it does not support,
  and what decision-makers may reasonably investigate or consider next.

The expectation chart is one tool in this process. It is not the product. The
product is the analyst's carefully qualified policy insight.

## Routine and non-random behavior

Most systems are influenced by many small forces. Measurements from such a
system fluctuate, even when no important change has occurred. An expectation
chart estimates a center and limits from the sequential behavior of the data.
Observations within those limits may look irregular while remaining consistent
with the routine system.

Some patterns warrant investigation—for example, an observation beyond an
expectation limit or a sufficiently long run on one side of the center. These
signals indicate that the pattern is difficult to explain as routine behavior
alone. They do not identify the responsible force and do not establish that a
particular policy caused the pattern.

## Prepare longitudinal data

```{r prepare}
policy_data <- data.frame(
  year = 2015:2024,
  measure = c(10, 11, 9, 12, 10, 11, 10, 19, 20, 21),
  jurisdiction = "Example Board"
)

prepared <- as_policy_data(policy_data, year, measure)
prepared[, c("date", "value", "jurisdiction")]
```

The explicit date and value selections are intentional. Real policy files can
contain multiple dates, targets, denominators, estimates, and subgroup values.
The analyst—not an automatic guess—should determine which fields answer the
question being considered.

## Calculate an untrended expectation chart

```{r calculate}
chart_data <- expectation_chart_data(policy_data, year, measure)
chart_data[, c(
  "date", "value", "center", "lower_limit", "upper_limit",
  "outside_limits"
)]
```

The function estimates sigma from the average moving range of successive
observations. Its limits describe the baseline represented by the supplied
data. They are not performance standards, desired targets, confidence
intervals, or acceptable-policy boundaries.

```{r plot, fig.width=7, fig.height=4.5}
plot(
  chart_data$date,
  chart_data$value,
  type = "b",
  xlab = "Year",
  ylab = "Measure",
  main = "Illustrative expectation chart"
)
lines(chart_data$date, chart_data$center, col = "steelblue", lwd = 2)
lines(chart_data$date, chart_data$upper_limit, col = "firebrick", lty = 2)
lines(chart_data$date, chart_data$lower_limit, col = "firebrick", lty = 2)
```

## Examine runs

```{r runs}
chart_data$run_signal <- detect_runs(chart_data$value, chart_data$center)
chart_data[, c("date", "value", "outside_limits", "run_signal")]
```

By default, `detect_runs()` marks the eighth and subsequent observations in a
run on the same side of the center. A value exactly on the center and a missing
value interrupt a run. Rules should be declared before interpreting the chart;
they should not be changed repeatedly until the analyst finds a preferred
story.

## Decide whether an expectation chart is suitable

Before relying on either a trended or untrended chart, ask:

1. Are observations comparable over time?
2. Did definitions, collection procedures, denominators, or reporting coverage
   change?
3. Is temporal spacing meaningful and correctly represented?
4. Does aggregation conceal materially different subgroups or jurisdictions?
5. Is a sustained trend substantively plausible?
6. Would a trended chart describe a stable direction, or merely normalize a
   policy-relevant change that should be investigated?

Run, line, bar, and cohort charts help the analyst inspect direction,
composition, gaps, and grouping structure before selecting an expectation-chart
form.

## Move from signal to insight

```{r summary}
longitudinal_summary(policy_data, year, measure)
```

The summary's `investigate` value is deliberately modest. Investigation may
include checking data quality, identifying changes in the system, consulting
program staff and affected communities, examining implementation timelines,
and testing alternative explanations.

A responsible policy briefing distinguishes:

- **Observation:** what the ordered measurements show.
- **Analytical reasoning:** why the pattern does or does not appear routine.
- **Context:** what institutional evidence and policy knowledge add.
- **Uncertainty:** what remains unknown or weakly supported.
- **Insight:** what decision-makers should understand or investigate.

The software supports the first two steps. The analyst is responsible for the
rest.
