---
title: "Using the policycraft Longitudinal Application"
author: "Dan Swart, CPA (ret)"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Using the policycraft Longitudinal Application}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
library(policycraft)
```

## Purpose and analytical responsibility

The application helps a policy analyst preserve temporal order, distinguish
routine variation from patterns worth investigating, and prepare evidence for
careful interpretation. A chart signal does not identify a cause or establish
that a policy worked. The analyst must combine the pattern with data quality,
institutional history, implementation evidence, affected-community knowledge,
and plausible alternative explanations.

Launch the locally installed application with:

```{r launch, eval=FALSE}
policycraft::launch_longitudinal()
```

Do not use `shiny::runApp("inst/app")` unless the working directory is the
package source repository. `launch_longitudinal()` locates the installed app.

## Choose an input

The upload control accepts CSV, Excel, data-frame RDS, and validated canonical
longitudinal RDS bundles. Ordinary files enter the mapping workflow. Canonical
bundles retain their registered observation grain, definitions, identifiers,
lineage, validation results, and screening warnings.

For ordinary files:

1. Select the column that orders the measurements.
2. Select the numeric value column.
3. Declare whether the ordering column contains **Calendar dates** or an
   **Observation sequence**.
4. Click **Apply mapping to app data**.
5. Confirm that the status begins with `Applied:` and inspect **Data Table**.

The proposed mapping does not alter the working data until it is applied. The
uploaded source file is never overwritten. A sequence such as `1:15` remains an
ordinal `observation` axis; it is not presented as dates in 1970.

## Filter and group deliberately

Filters restrict the rows shared by the analytical tabs. The grouping selector
creates distinct lines or bars from an existing column; it does not aggregate
unrelated series into a new measure. For a canonical bundle, Run, expectation,
and autocorrelation diagnostics require exactly one existing `series_id`.

Before interpreting a chart, confirm the measure definition, unit, reporting
status, comparable coverage, missing periods, and any changes in collection or
calculation. Missing periods remain gaps and interrupt moving-range and run
calculations.

## Read the chart tabs progressively

### Run Chart

Use the Run Chart as the first view of ordered behavior. It connects the values
and adds the median. Look for direction, clustering, gaps, and observations that
deserve verification before using expectation limits.

### Line Chart

Use the Line Chart for one series or for an explicit comparison selected in
**Line/Bar Chart Grouping**. Grouped lines receive endpoint labels. A comparison
can reveal composition or divergent trajectories, but does not itself explain
the differences.

### Bar Chart

The Bar Chart summarizes the currently selected values by the chosen grouping
column. Use it for comparison, not as a substitute for the ordered evidence.

### Untrended Expectation Chart

This chart estimates a center line (CL) and upper and lower expectation limits
(UCL and LCL) from successive moving ranges. Limits describe the system
represented by the selected observations; they are not targets, confidence
intervals, or acceptable-performance standards.

The chart reports:

- observations outside UCL or LCL;
- eight consecutive observations on one side of CL;
- six consecutive observations steadily increasing or decreasing; and
- fourteen consecutive observations alternating direction.

Only the endpoint at which a rule becomes satisfied and later qualifying
endpoints are flagged. Ties, missing values, and recalculation boundaries
interrupt the applicable sequence.

Enable recalculation when there is a defensible intervention or system-change
boundary. Select a date for calendar data or an observation number for ordinal
data. Observations before the boundary establish the original frozen limits;
observations beginning at the boundary establish the recalculated segment.

### Trended Expectation Chart

Use a trended chart only when a sustained direction is a plausible routine
feature of the system. The fitted linear trend becomes the center line and the
limits follow it. For observation sequences, elapsed position—not artificial
days—is used. A trend model can normalize a meaningful change, so compare it
with the untrended chart and explain why the trended form is appropriate.

### Cohort and autocorrelation views

The Cohort Chart follows a selected group through successive grades and years.
The autocorrelation tab reports lagged relationships within the ordered series.
Autocorrelation can affect limit estimation, but it is not evidence that one
policy variable caused another.

### Runs Debug

Runs Debug exposes the separate same-side, monotonic-trend, alternating, and
combined flags. Use it to verify which observation triggered a rule rather than
to search repeatedly for a preferred signal.

## Export charts and analysis-ready code

PNG is convenient for routine documents, SVG for scalable editing, and PDF for
print workflows. **Download R Code** creates a compact `.R` script that can be
opened in RStudio, sourced, or pasted into a Quarto code cell. It includes:

- a recognizable `chart_data` data frame;
- named calculations used by the chart;
- ordinary `ggplot()`, `aes()`, and `geom_*()` calls;
- the visible axis graduations;
- editable CL, UCL, and LCL annotations; and
- the shared `policycraft_chart_theme()` call.

For example:

```{r exported-code, eval=FALSE}
source("expectation_chart_2026-09-04.R")
chart

# Continue editing as a normal ggplot object.
chart + ggplot2::labs(
  title = "Revised analytical title",
  subtitle = "Define the series and period",
  caption = "Source and interpretive caution"
)
```

The generated expectation-chart limit labels default to bold text with
`size = 4`. Major and minor panel grids are removed. These are ordinary ggplot2
arguments and may be changed in the downloaded script.

## A disciplined completion checklist

Before presenting a result, confirm that:

1. the ordering and value mappings are correct and applied;
2. filters leave the intended analytical series and period;
3. dates or observation numbers have the intended meaning;
4. missingness and definition changes are disclosed;
5. the selected chart form is justified;
6. run-rule and limit signals are described as reasons to investigate;
7. titles, units, source, and captions make the exported chart self-explanatory;
8. causal language is supported by evidence beyond the chart; and
9. the final QMD retains the editable code used to generate the figure.

